---
title: "Aim 2 — Variable Creation"
author: "Tianqi Ouyang"
date: "2026-06-30"
---

This file builds every analysis variable specified in the Aim 2 Statistical
Analysis Plan (`Tianqi_SAP.docx`) from two data sources and writes one row per
participant (`aim2_master.rds`) for the downstream Table 1 / analysis files.

Variables are grouped by **data source**:

- **REDCap** — the longitudinal study database (demographics, comorbidities,
  kidney function, ICI regimen, concomitant medications).
- **RPDR** — the hospital laboratory pull (baseline hemoglobin and albumin).

Each table below lists the variable, its type, and its definition. The R code
that creates it is hidden — click **► Code** to expand any block.

> **REDCap is longitudinal.** The export has one row per
> (`record_id` × `redcap_event_name`); each variable is pulled from its event,
> then joined back to one row per `record_id`:
>
> | Event (`redcap_event_name`) | Variables sourced here |
> |---|---|
> | `prebaseline_arm_1` | demographics, comorbidities, concomitant meds |
> | `baseline_arm_1` | `egfr`, baseline `visit_date`, `EMPI` |
> | `early_treatment_we_arm_1` | `ici_type` (baseline ICI regimen) |

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(readxl)
library(lubridate)
```

```{r}
#| label: config
# --- FILL IN once the Aim 2 data arrives (logic below is fully SAP-specified) ---
redcap_path   <- "Data/AIM2_REDCAP_DATA.csv"   # longitudinal REDCap export
rpdr_lab_path <- "Data/AIM2_RPDR_Lab.txt"       # RPDR lab pull (pipe-delimited), keyed by EMPI
HGB_GROUP     <- "HGB"   # hemoglobin Group_Id in the RPDR pull
ALB_GROUP     <- "ALB"   # albumin   Group_Id in the RPDR pull
```

```{r}
#| label: load-redcap
raw <- read_csv(redcap_path, show_col_types = FALSE)
stopifnot("redcap_event_name" %in% names(raw))

prebaseline <- raw %>% filter(redcap_event_name == "prebaseline_arm_1")
baseline    <- raw %>% filter(redcap_event_name == "baseline_arm_1")
early_tx    <- raw %>% filter(redcap_event_name == "early_treatment_we_arm_1")

# Helper: OR across one or more 0/1 checkbox columns → 0/1 (NA treated as 0)
any_box <- function(df, cols) {
  as.integer(rowSums(across(all_of(cols), ~ replace_na(.x == 1, FALSE))) > 0)
}
```

# REDCap

| Variable | Type | Definition (REDCap source fields) |
|---|---|---|
| `sex`, `race`, `ethnicity`, `study_arm` | factor | REDCap `sex`, `race`, `ethnicity`, `study_arm` (pre-baseline), labelled as factors |
| `cirrhosis` | 0/1 | REDCap `comorb_gi___9` |
| `asvcd` | 0/1 | REDCap `comorb_cardiac___1` OR `comorb_neuro___3` |
| `chf` | 0/1 | REDCap `comorb_cardiac___5` |
| `hypertension` | 0/1 | REDCap `comorb_cardiac___4` |
| `copd` | 0/1 | REDCap `comorb_pulm___2` |
| `diabetes` | 0/1 | REDCap `comorb_endocrine___3` OR `comorb_endocrine___4` |
| `autoimmune_history` | 0/1 | 1 if **any** of 18 REDCap boxes checked: `comorb_rheum___1`, `comorb_rheum___3`, `comorb_pulm___3`, `comorb_cardiac___6`, `comorb_gi___1`, `comorb_gi___3`, `comorb_gi___4`, `comorb_gi___5`, `comorb_gi___8`, `comorb_derm___1`, `comorb_derm___2`, `comorb_derm___3`, `comorb_derm___4`, `comorb_neuro___1`, `comorb_neuro___2`, `comorb_endocrine___2`, `comorb_endocrine___3`, `comorb_heme___1` |
| `baseline_egfr` | continuous | REDCap `egfr` at baseline event (mL/min/1.73 m²) |
| `baseline_ckd_stage` | factor G1–G5 | KDIGO GFR categories of REDCap `egfr` |
| `baseline_ckd_binary` | 0/1 | 1 if REDCap `egfr` <60 (G3a–G5) |
| `baseline_ici_treatment` | factor (11) | REDCap `ici_type` at early-treatment event |
| `ici_regimen_class` | factor (4) | mechanism class of REDCap `ici_type` |
| `ppi`, `antibiotic`, `tmp_smx`, `nsaid`, `allopurinol`, `antiepileptic` | 0/1 | structured REDCap field — `ppi`, `nephrotox_drugs___1` (antibiotic), `___2` (tmp_smx), `___3` (nsaid), `___4` (allopurinol), `___5` (antiepileptic) — OR a `con_med` free-text match |
| `AIN_con_med_any` | 0/1 | 1 if any of the six medication classes above is present |

```{r}
#| label: demographics
# Demographics — numeric REDCap codes labelled as factors for Table 1.
demographics <- prebaseline %>%
  transmute(
    record_id,
    sex = factor(sex, levels = c(1, 2, 3),
                 labels = c("Male", "Female", "Other")),
    race = factor(race, levels = c(1, 2, 3, 4),
                  labels = c("White", "Black", "Asian", "Other")),
    ethnicity = factor(ethnicity, levels = c(1, 2),
                       labels = c("Hispanic", "Non-Hispanic")),
    study_arm = factor(
      study_arm, levels = c(1, 2, 3),
      labels = c("Metastatic melanoma + ICI",
                 "Adjuvant stage II–III melanoma + ICI",
                 "Control stage I–IIIA melanoma, no ICI"))
  )
```

```{r}
#| label: comorbidities
# Binary / composite comorbidities (0 = Absent, 1 = Present).
comorbidities <- prebaseline %>%
  transmute(
    record_id,
    cirrhosis    = any_box(prebaseline, "comorb_gi___9"),
    asvcd        = any_box(prebaseline, c("comorb_cardiac___1", "comorb_neuro___3")),
    chf          = any_box(prebaseline, "comorb_cardiac___5"),
    hypertension = any_box(prebaseline, "comorb_cardiac___4"),
    copd         = any_box(prebaseline, "comorb_pulm___2"),
    diabetes     = any_box(prebaseline, c("comorb_endocrine___3", "comorb_endocrine___4"))
  )
```

```{r}
#| label: autoimmune-history
# autoimmune_history = 1 if ANY pre-existing autoimmune comorbidity box is checked.
autoimmune_boxes <- c(
  "comorb_rheum___1", "comorb_rheum___3",
  "comorb_pulm___3",
  "comorb_cardiac___6",
  "comorb_gi___1", "comorb_gi___3", "comorb_gi___4", "comorb_gi___5", "comorb_gi___8",
  "comorb_derm___1", "comorb_derm___2", "comorb_derm___3", "comorb_derm___4",
  "comorb_neuro___1", "comorb_neuro___2",
  "comorb_endocrine___2", "comorb_endocrine___3",
  "comorb_heme___1"
)

autoimmune <- prebaseline %>%
  transmute(record_id,
            autoimmune_history = any_box(prebaseline, autoimmune_boxes))
```

```{r}
#| label: ckd
# Baseline kidney function: egfr at baseline event → KDIGO stage + <60 binary.
# KDIGO bands are inclusive at the top (G2 = 60–89, G1 ≥90); right = FALSE
# with these breaks reproduces that: <15 G5 · 15–29 G4 · 30–44 G3b ·
# 45–59 G3a · 60–89 G2 · ≥90 G1.
kidney <- baseline %>%
  transmute(
    record_id,
    baseline_egfr = as.numeric(egfr),
    baseline_ckd_stage = cut(
      baseline_egfr,
      breaks = c(-Inf, 15, 30, 45, 60, 90, Inf),
      labels = c("G5", "G4", "G3b", "G3a", "G2", "G1"),
      right  = FALSE),
    baseline_ckd_binary = if_else(baseline_egfr < 60, 1L, 0L)  # 1 = G3a–G5
  ) %>%
  mutate(baseline_ckd_stage = factor(
    baseline_ckd_stage, levels = c("G1", "G2", "G3a", "G3b", "G4", "G5")))
```

```{r}
#| label: ici
# Baseline ICI regimen from ici_type at the Early Treatment (Week 6–12) event.
# ici_regimen_class collapses the 11 types into 4 mechanism classes.
ici <- early_tx %>%
  transmute(
    record_id,
    baseline_ici_treatment = factor(
      ici_type, levels = 1:11,
      labels = c("Pembrolizumab", "Nivolumab", "Ipilimumab", "Relatlimab",
                 "Avelumab", "Durvalumab", "Atezolizumab",
                 "Nivolumab/Ipilimumab", "Nivolumab/Relatlimab",
                 "Other", "Not receiving ICI")),
    ici_regimen_class = factor(
      case_when(
        ici_type %in% c(1, 2, 5, 6, 7) ~ 1L,   # PD-(L)1 monotherapy
        ici_type %in% c(3, 8)          ~ 2L,   # CTLA-4–containing
        ici_type %in% c(4, 9)          ~ 3L,   # LAG-3–containing
        ici_type %in% c(10, 11)        ~ 4L    # Other / No active ICI
      ),
      levels = 1:4,
      labels = c("PD-(L)1 monotherapy", "CTLA-4–containing",
                 "LAG-3–containing", "Other/No active ICI"))
  )
```

```{r}
#| label: con-med-search-terms
# Concomitant meds: six AIN-associated classes. Each free-text term is matched
# case-insensitively against con_med; a class is present if its structured
# REDCap field == 1 OR a term matches.
search_terms <- list(
  ppi = c("omeprazole","Prilosec","esomeprazole","Nexium","pantoprazole","Protonix",
          "lansoprazole","Prevacid","dexlansoprazole","Dexilant","rabeprazole","Aciphex"),
  antibiotic = c("amoxicillin","Augmentin","penicillin","ampicillin","dicloxacillin",
          "cephalexin","Keflex","cefazolin","ceftriaxone","Rocephin","cefepime",
          "cefpodoxime","cefdinir","ciprofloxacin","Cipro","levofloxacin","Levaquin",
          "moxifloxacin","Avelox","azithromycin","Zithromax","clarithromycin","Biaxin",
          "erythromycin","doxycycline","minocycline","tetracycline","clindamycin",
          "Cleocin","metronidazole","Flagyl","nitrofurantoin","Macrobid","fosfomycin",
          "Monurol","vancomycin","linezolid","Zyvox"),
  tmp_smx = c("trimethoprim-sulfamethoxazole","TMP-SMX","TMP/SMX","sulfamethoxazole",
          "cotrimoxazole","co-trimoxazole","Bactrim","Septra"),
  nsaid = c("ibuprofen","Advil","Motrin","naproxen","Aleve","diclofenac","Voltaren",
          "indomethacin","Indocin","meloxicam","Mobic","celecoxib","Celebrex",
          "ketorolac","Toradol","nabumetone","Relafen","sulindac","etodolac",
          "piroxicam","Feldene","oxaprozin","Daypro"),
  allopurinol = c("allopurinol","Zyloprim","Aloprim"),
  antiepileptic = c("carbamazepine","Tegretol","oxcarbazepine","Trileptal","phenytoin",
          "Dilantin","fosphenytoin","Cerebyx","lamotrigine","Lamictal","phenobarbital")
)

structured_field <- c(
  ppi           = "ppi",
  antibiotic    = "nephrotox_drugs___1",
  tmp_smx       = "nephrotox_drugs___2",
  nsaid         = "nephrotox_drugs___3",
  allopurinol   = "nephrotox_drugs___4",
  antiepileptic = "nephrotox_drugs___5"
)
```

```{r}
#| label: con-med
# Build a case-insensitive alternation regex per class (terms regex-escaped),
# then flag each class and AIN_con_med_any.
esc <- function(x) gsub("([][(){}.^$*+?\\\\|-])", "\\\\\\1", x)
class_regex <- map_chr(search_terms, ~ paste0(esc(.x), collapse = "|"))

con_text <- prebaseline$con_med %>% replace_na("")

med_flags <- imap_dfc(structured_field, function(field, cls) {
  structured <- if (field %in% names(prebaseline)) {
    replace_na(prebaseline[[field]] == 1, FALSE)
  } else FALSE
  freetext <- str_detect(con_text, regex(class_regex[[cls]], ignore_case = TRUE))
  tibble(!!cls := as.integer(structured | freetext))
})

medications <- bind_cols(record_id = prebaseline$record_id, med_flags) %>%
  mutate(AIN_con_med_any = as.integer(
    rowSums(across(all_of(names(structured_field)))) > 0))
```

# RPDR

| Variable | Type | Definition (RPDR value, anchored on REDCap fields) |
|---|---|---|
| `baseline_hgb` | continuous | RPDR hemoglobin (`Group_Id` = `HGB`) closest to the baseline visit, anchored on REDCap `visit_date` and joined via REDCap `EMPI` (±30 d; ties prefer prior) |
| `baseline_albumin` | continuous | RPDR albumin (`Group_Id` = `ALB`) closest to the baseline visit, anchored on REDCap `visit_date` and joined via REDCap `EMPI` (±30 d; ties prefer prior) |

> Requires an `EMPI` ↔ `record_id` crosswalk so RPDR (keyed by `EMPI`) can join
> to REDCap (keyed by `record_id`). This block is `eval: false` until the RPDR
> pull and crosswalk are available.

```{r}
#| label: rpdr-labs
#| eval: false
# Closest-lab-to-date: smallest |day-diff| from the baseline visit; ties prefer
# the prior value (diff <= 0); anything > 30 days away is coded missing.
closest_lab <- function(lab_df, index_df, group_id, varname, window = 30) {
  lab_df %>%
    filter(Group_Id == group_id, grepl("^\\d+\\.?\\d*$", Result)) %>%
    mutate(Result = as.numeric(Result)) %>%
    inner_join(index_df, by = "EMPI") %>%
    mutate(diff = as.numeric(Seq_Date_Time - baseline_visit_date)) %>%
    filter(abs(diff) <= window) %>%
    arrange(EMPI, abs(diff), diff > 0) %>%
    distinct(EMPI, .keep_all = TRUE) %>%
    transmute(EMPI, !!varname := Result)
}

lab <- read.table(rpdr_lab_path, sep = "|", fill = TRUE, quote = "",
                  header = TRUE) %>%
  select(EMPI, Seq_Date_Time, Result, Group_Id)
lab$Seq_Date_Time <- as.Date(lab$Seq_Date_Time, "%m/%d/%Y")

index_df <- baseline %>%
  transmute(record_id, EMPI, baseline_visit_date = as.Date(visit_date)) %>%
  filter(!is.na(EMPI))

hgb <- closest_lab(lab, index_df, HGB_GROUP, "baseline_hgb")
alb <- closest_lab(lab, index_df, ALB_GROUP, "baseline_albumin")

rpdr_labs <- index_df %>%
  select(record_id, EMPI) %>%
  left_join(hgb, by = "EMPI") %>%
  left_join(alb, by = "EMPI") %>%
  select(record_id, baseline_hgb, baseline_albumin)
```

```{r}
#| label: assemble
# Assemble one row per participant and write aim2_master.rds. RPDR labs are
# attached only if that block was run (data available).
aim2_master <- demographics %>%
  left_join(comorbidities, by = "record_id") %>%
  left_join(autoimmune,    by = "record_id") %>%
  left_join(kidney,        by = "record_id") %>%
  left_join(ici,           by = "record_id") %>%
  left_join(medications,   by = "record_id")

if (exists("rpdr_labs")) {
  aim2_master <- aim2_master %>% left_join(rpdr_labs, by = "record_id")
}

saveRDS(aim2_master, "aim2_master.rds")
glimpse(aim2_master)
```
