---
title: "Aim 2 — Variable Creation"
author: "Tianqi Ouyang"
date: "2026-06-30"
---

# 0 Purpose

This file builds every **analysis variable specified in the Aim 2 Statistical
Analysis Plan** (`Tianqi_SAP.docx`) from the longitudinal REDCap export and the
RPDR laboratory pull. It produces one row per participant (`aim2_master.rds`)
that the downstream Table 1 / analysis files consume.

Aim 2 is a melanoma + immune-checkpoint-inhibitor (ICI) cohort examining
acute interstitial nephritis (AIN). Study arms:

1. Metastatic melanoma receiving ICI
2. Adjuvant stage II–III melanoma receiving ICI
3. Control stage I–IIIA melanoma **not** receiving ICI

> **REDCap structure.** This is a *longitudinal* project. The export has one
> row per (`record_id` × `redcap_event_name`). Variables live in specific
> events, so each block below pulls its variables from the correct event and
> the blocks are joined back to one row per `record_id`:
>
> | Event (`redcap_event_name`) | Variables sourced here |
> |---|---|
> | `prebaseline_arm_1` | demographics, comorbidities, concomitant meds (`ppi`, `nephrotox_drugs___*`, `con_med`) |
> | `baseline_arm_1` | `egfr`, baseline `visit_date` |
> | `early_treatment_we_arm_1` | `ici_type` (defines baseline ICI regimen) |

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(readxl)
library(lubridate)
```

# 1 Configuration — input files

> **FILL IN once the Aim 2 data arrives.** The variable logic below is fully
> specified by the SAP; only these paths are project-specific.

```{r}
#| label: config
# Longitudinal REDCap export for the melanoma/ICI/AIN project
redcap_path <- "Data/AIM2_REDCAP_DATA.csv"          # <-- set me

# RPDR laboratory pull (pipe-delimited .txt), keyed by EMPI.
# Needed only for baseline_hgb / baseline_albumin (closest lab to baseline visit).
rpdr_lab_path <- "Data/AIM2_RPDR_Lab.txt"           # <-- set me

# RPDR Group_Id labels for the two labs we need from RPDR
HGB_GROUP <- "HGB"   # hemoglobin Group_Id in the RPDR pull
ALB_GROUP <- "ALB"   # albumin   Group_Id in the RPDR pull
```

# 2 Load REDCap & split by event

```{r}
#| label: load-redcap
raw <- read_csv(redcap_path, show_col_types = FALSE)

# Defensive: REDCap exports the event column as redcap_event_name.
stopifnot("redcap_event_name" %in% names(raw))

prebaseline <- raw %>% filter(redcap_event_name == "prebaseline_arm_1")
baseline    <- raw %>% filter(redcap_event_name == "baseline_arm_1")
early_tx    <- raw %>% filter(redcap_event_name == "early_treatment_we_arm_1")
```

# 3 Demographics (Pre-baseline)

Stored as numeric REDCap codes; labelled here as factors for Table 1.

```{r}
#| label: demographics
demographics <- prebaseline %>%
  transmute(
    record_id,
    sex = factor(sex, levels = c(1, 2, 3),
                 labels = c("Male", "Female", "Other")),
    race = factor(race, levels = c(1, 2, 3, 4),
                  labels = c("White", "Black", "Asian", "Other")),
    ethnicity = factor(ethnicity, levels = c(1, 2),
                       labels = c("Hispanic", "Non-Hispanic")),
    study_arm = factor(
      study_arm, levels = c(1, 2, 3),
      labels = c("Metastatic melanoma + ICI",
                 "Adjuvant stage II–III melanoma + ICI",
                 "Control stage I–IIIA melanoma, no ICI"))
  )
```

# 4 Baseline comorbidities (Pre-baseline)

REDCap checkbox fields export as `field___N` coded 0/1. Composite comorbidities
combine the relevant boxes with logical OR; binary comorbidities map a single
box. All coded 0 = Absent, 1 = Present.

```{r}
#| label: comorbidities
# Helper: OR across one or more 0/1 checkbox columns → 0/1 (NA treated as 0)
any_box <- function(df, cols) {
  as.integer(rowSums(across(all_of(cols), ~ replace_na(.x == 1, FALSE))) > 0)
}

comorbidities <- prebaseline %>%
  transmute(
    record_id,
    cirrhosis    = any_box(prebaseline, "comorb_gi___9"),
    asvcd        = any_box(prebaseline, c("comorb_cardiac___1", "comorb_neuro___3")),
    chf          = any_box(prebaseline, "comorb_cardiac___5"),
    hypertension = any_box(prebaseline, "comorb_cardiac___4"),
    copd         = any_box(prebaseline, "comorb_pulm___2"),
    diabetes     = any_box(prebaseline, c("comorb_endocrine___3", "comorb_endocrine___4"))
  )
```

## 4.1 Derived: Autoimmune history (`autoimmune_history`)

`autoimmune_history` = 1 if **any** pre-existing autoimmune comorbidity box is
checked, else 0 (SAP §"Derived variable: Autoimmune history").

```{r}
#| label: autoimmune-history
autoimmune_boxes <- c(
  "comorb_rheum___1", "comorb_rheum___3",
  "comorb_pulm___3",
  "comorb_cardiac___6",
  "comorb_gi___1", "comorb_gi___3", "comorb_gi___4", "comorb_gi___5", "comorb_gi___8",
  "comorb_derm___1", "comorb_derm___2", "comorb_derm___3", "comorb_derm___4",
  "comorb_neuro___1", "comorb_neuro___2",
  "comorb_endocrine___2", "comorb_endocrine___3",
  "comorb_heme___1"
)

autoimmune <- prebaseline %>%
  transmute(record_id,
            autoimmune_history = any_box(prebaseline, autoimmune_boxes))
```

# 5 Baseline kidney function (Baseline event)

`egfr` at `baseline_arm_1` (mL/min/1.73 m²). CKD stage by KDIGO GFR categories;
plus the prespecified collapsed binary (<60 vs ≥60).

```{r}
#| label: ckd
kidney <- baseline %>%
  transmute(
    record_id,
    baseline_egfr = as.numeric(egfr),
    baseline_ckd_stage = cut(
      baseline_egfr,
      breaks = c(-Inf, 15, 30, 45, 60, 90, Inf),
      labels = c("G5", "G4", "G3b", "G3a", "G2", "G1"),
      right  = FALSE),                              # [lower, upper)
    # KDIGO boundaries are inclusive at the top of each band (e.g. G2 = 60–89,
    # G1 ≥90). right = FALSE with these breaks reproduces that exactly:
    #   <15 G5 · 15–29 G4 · 30–44 G3b · 45–59 G3a · 60–89 G2 · ≥90 G1
    baseline_ckd_binary = if_else(baseline_egfr < 60, 1L, 0L)  # 1 = G3a–G5
  ) %>%
  mutate(baseline_ckd_stage = factor(
    baseline_ckd_stage, levels = c("G1", "G2", "G3a", "G3b", "G4", "G5")))
```

# 6 Baseline ICI treatment (Early Treatment event)

Baseline ICI regimen is defined by `ici_type` recorded at the **Early Treatment
(Week 6–12)** event — the most accurate reflection of the established regimen
(SAP §"Derived variable: Baseline ICI Treatment"). `ici_regimen_class` collapses
the 11 types into 4 mechanism classes.

```{r}
#| label: ici
ici <- early_tx %>%
  transmute(
    record_id,
    baseline_ici_treatment = factor(
      ici_type, levels = 1:11,
      labels = c("Pembrolizumab", "Nivolumab", "Ipilimumab", "Relatlimab",
                 "Avelumab", "Durvalumab", "Atezolizumab",
                 "Nivolumab/Ipilimumab", "Nivolumab/Relatlimab",
                 "Other", "Not receiving ICI")),
    ici_regimen_class = factor(
      case_when(
        ici_type %in% c(1, 2, 5, 6, 7) ~ 1L,   # PD-(L)1 monotherapy
        ici_type %in% c(3, 8)          ~ 2L,   # CTLA-4–containing
        ici_type %in% c(4, 9)          ~ 3L,   # LAG-3–containing
        ici_type %in% c(10, 11)        ~ 4L    # Other / No active ICI
      ),
      levels = 1:4,
      labels = c("PD-(L)1 monotherapy", "CTLA-4–containing",
                 "LAG-3–containing", "Other/No active ICI"))
  )
```

# 7 Baseline concomitant medications (Pre-baseline)

Six AIN-associated medication classes. Exposure = present (1) if the **structured
REDCap field equals 1 OR** a generic/brand search term is found in the free-text
`con_med` field (case-insensitive). Then `AIN_con_med_any` = 1 if any of the six
classes is present.

```{r}
#| label: con-med-search-terms
# Free-text search terms per class (SAP medication-class table). Each term is
# matched case-insensitively as a word-boundary regex against con_med.
search_terms <- list(
  ppi = c("omeprazole","Prilosec","esomeprazole","Nexium","pantoprazole","Protonix",
          "lansoprazole","Prevacid","dexlansoprazole","Dexilant","rabeprazole","Aciphex"),
  antibiotic = c("amoxicillin","Augmentin","penicillin","ampicillin","dicloxacillin",
          "cephalexin","Keflex","cefazolin","ceftriaxone","Rocephin","cefepime",
          "cefpodoxime","cefdinir","ciprofloxacin","Cipro","levofloxacin","Levaquin",
          "moxifloxacin","Avelox","azithromycin","Zithromax","clarithromycin","Biaxin",
          "erythromycin","doxycycline","minocycline","tetracycline","clindamycin",
          "Cleocin","metronidazole","Flagyl","nitrofurantoin","Macrobid","fosfomycin",
          "Monurol","vancomycin","linezolid","Zyvox"),
  tmp_smx = c("trimethoprim-sulfamethoxazole","TMP-SMX","TMP/SMX","sulfamethoxazole",
          "cotrimoxazole","co-trimoxazole","Bactrim","Septra"),
  nsaid = c("ibuprofen","Advil","Motrin","naproxen","Aleve","diclofenac","Voltaren",
          "indomethacin","Indocin","meloxicam","Mobic","celecoxib","Celebrex",
          "ketorolac","Toradol","nabumetone","Relafen","sulindac","etodolac",
          "piroxicam","Feldene","oxaprozin","Daypro"),
  allopurinol = c("allopurinol","Zyloprim","Aloprim"),
  antiepileptic = c("carbamazepine","Tegretol","oxcarbazepine","Trileptal","phenytoin",
          "Dilantin","fosphenytoin","Cerebyx","lamotrigine","Lamictal","phenobarbital")
)

# Structured REDCap field backing each class
structured_field <- c(
  ppi           = "ppi",
  antibiotic    = "nephrotox_drugs___1",
  tmp_smx       = "nephrotox_drugs___2",
  nsaid         = "nephrotox_drugs___3",
  allopurinol   = "nephrotox_drugs___4",
  antiepileptic = "nephrotox_drugs___5"
)
```

```{r}
#| label: con-med
# Build a case-insensitive alternation regex per class (terms regex-escaped).
esc <- function(x) gsub("([][(){}.^$*+?\\\\|-])", "\\\\\\1", x)
class_regex <- map_chr(search_terms, ~ paste0(esc(.x), collapse = "|"))

con_text <- prebaseline$con_med %>% replace_na("")

med_flags <- imap_dfc(structured_field, function(field, cls) {
  structured <- if (field %in% names(prebaseline)) {
    replace_na(prebaseline[[field]] == 1, FALSE)
  } else FALSE
  freetext <- str_detect(con_text, regex(class_regex[[cls]], ignore_case = TRUE))
  tibble(!!cls := as.integer(structured | freetext))
})

medications <- bind_cols(record_id = prebaseline$record_id, med_flags) %>%
  mutate(AIN_con_med_any = as.integer(
    rowSums(across(all_of(names(structured_field)))) > 0))
```

# 8 Baseline RPDR labs (closest to baseline visit)

`baseline_hgb` and `baseline_albumin` are the RPDR results with the **smallest
absolute day-difference** from the baseline visit date. Ties prefer the value
obtained **prior to** the visit. Anything > 30 days away is coded missing
(SAP §"Baseline Hemoglobin / Albumin").

> Requires an `EMPI` ↔ `record_id` crosswalk so RPDR (keyed by EMPI) can join to
> REDCap (keyed by record_id). Set `EMPI` from the export if present.

```{r}
#| label: rpdr-labs
#| eval: false
# Closest-lab-to-date helper: smallest |day-diff| from the index date.
closest_lab <- function(lab_df, index_df, group_id, varname, window = 30) {
  lab_df %>%
    filter(Group_Id == group_id, grepl("^\\d+\\.?\\d*$", Result)) %>%
    mutate(Result = as.numeric(Result)) %>%
    inner_join(index_df, by = "EMPI") %>%
    mutate(diff = as.numeric(Seq_Date_Time - baseline_visit_date)) %>%
    filter(abs(diff) <= window) %>%
    # smallest |diff| first; ties → prior (diff <= 0) preferred
    arrange(EMPI, abs(diff), diff > 0) %>%
    distinct(EMPI, .keep_all = TRUE) %>%
    transmute(EMPI, !!varname := Result)
}

lab <- read.table(rpdr_lab_path, sep = "|", fill = TRUE, quote = "",
                  header = TRUE) %>%
  select(EMPI, Seq_Date_Time, Result, Group_Id)
lab$Seq_Date_Time <- as.Date(lab$Seq_Date_Time, "%m/%d/%Y")

# baseline visit date + EMPI per record (visit_date lives in baseline_arm_1)
index_df <- baseline %>%
  transmute(record_id, EMPI, baseline_visit_date = as.Date(visit_date)) %>%
  filter(!is.na(EMPI))

hgb <- closest_lab(lab, index_df, HGB_GROUP, "baseline_hgb")
alb <- closest_lab(lab, index_df, ALB_GROUP, "baseline_albumin")

rpdr_labs <- index_df %>%
  select(record_id, EMPI) %>%
  left_join(hgb, by = "EMPI") %>%
  left_join(alb, by = "EMPI") %>%
  select(record_id, baseline_hgb, baseline_albumin)
```

# 9 Assemble one row per participant

```{r}
#| label: assemble
aim2_master <- demographics %>%
  left_join(comorbidities, by = "record_id") %>%
  left_join(autoimmune,    by = "record_id") %>%
  left_join(kidney,        by = "record_id") %>%
  left_join(ici,           by = "record_id") %>%
  left_join(medications,   by = "record_id")

# Attach RPDR labs only if that chunk was run (data available)
if (exists("rpdr_labs")) {
  aim2_master <- aim2_master %>% left_join(rpdr_labs, by = "record_id")
}

saveRDS(aim2_master, "aim2_master.rds")
glimpse(aim2_master)
```

# 10 Variable dictionary (created here)

| Variable | Type | Source / definition |
|---|---|---|
| `sex`, `race`, `ethnicity`, `study_arm` | factor | Pre-baseline demographics (labelled) |
| `cirrhosis` | 0/1 | `comorb_gi___9` |
| `asvcd` | 0/1 | `comorb_cardiac___1` OR `comorb_neuro___3` |
| `chf` | 0/1 | `comorb_cardiac___5` |
| `hypertension` | 0/1 | `comorb_cardiac___4` |
| `copd` | 0/1 | `comorb_pulm___2` |
| `diabetes` | 0/1 | `comorb_endocrine___3` OR `comorb_endocrine___4` |
| `autoimmune_history` | 0/1 | any of 18 autoimmune comorbidity boxes |
| `baseline_egfr` | continuous | `egfr` @ baseline |
| `baseline_ckd_stage` | factor G1–G5 | KDIGO categories of `baseline_egfr` |
| `baseline_ckd_binary` | 0/1 | 1 if eGFR <60 (G3a–G5) |
| `baseline_ici_treatment` | factor (11) | `ici_type` @ early-treatment |
| `ici_regimen_class` | factor (4) | mechanism class of `ici_type` |
| `ppi`,`antibiotic`,`tmp_smx`,`nsaid`,`allopurinol`,`antiepileptic` | 0/1 | structured field OR `con_med` free-text |
| `AIN_con_med_any` | 0/1 | any of the six medication classes |
| `baseline_hgb`, `baseline_albumin` | continuous | RPDR lab closest to baseline visit (±30 d) |
